IALE 2015 Analysis

### 2015-07-03 iale2015_sem.Rmd
Calculated plot level leaf count variables, joined plot-level environmental; calculated understory richness from vegetation transects, and overstory richness from tagged and untagged stem DBH data and saved this into a "richness" data frame. Calculated multiple DBH variables for tagged & untagged stems, combined those data frames together, and calculated relative density, relative dominance, and importance value for each species.

```{r load data}
# Ecological data: leaf counts, infection
load("~/GitHub/superspreaders/stems_plots.RData")

# Physical environment: topography
library(rgdal)
plots_env <- readOGR("analysis/data/plot_gis/", "soco_plots_metric")
str(plots_env)
plot(plots_env, axes=T)
plots_env <- plots_env@data

# Temperature: rainy season hour summaries
plots_temps <- read.csv("analysis/data/rainy_season_temps.csv")
summary(plots_temps)
plots_temps <- select(plots_temps, -X)
```

Join the plot level environmental data (geographic/topographic & temperatures)
```{r join plot data, message=FALSE}
library(dplyr)
plots_env <- left_join(plots_env, plots_temps, by = c("PLOT_ID"="plotid"))
summary(plots_env)
str(plots_env)
plots_env <- plots_env %>% select(-cat) %>% rename(easting = X, northing = Y, plotid = PLOT_ID)
```

Investigating `NAs` in the data set
```{r investigate NAs}
summary(plots_env)
filter(plots_env, is.na(sample_year))

@@

Make a pairs plot of the continuous variables, i.e. everything except `plotid`.
```{r pairs plot}
# Helper functions for `pairs`
panel.hist <- function(x, ...)
{
      usr <- par("usr"); on.exit(par(usr))
      par(usr = c(usr[1:2], 0, 1.5) )
      h <- hist(x, plot = FALSE)
      breaks <- h$breaks; nB <- length(breaks)
      y <- h$counts; y <- y/max(y)
      rect(breaks[-nB], 0, breaks[-1], y, col = "cyan", ...)
}
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
      usr <- par("usr"); on.exit(par(usr))
      par(usr = c(0, 1, 0, 1))
      r <- abs(cor(x, y, use = "na.or.complete"))
      txt <- format(c(r, 0.123456789), digits = digits)[1]
      txt <- paste0(prefix, txt)
      if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
      text(0.5, 0.5, txt, cex = cex.cor * r)
}

pairs(select(plots_env, -plotid, -sample_year), lower.panel = panel.cor, diag.panel = panel.hist)

plots_env$avg_psi <- rowMeans(cbind(plots_env$psiSpr, plots_env$psiSum, plots_env$psiWin))
pairs(select(plots_env, -plotid, -sample_year, -starts_with("psi")), lower.panel = panel.cor, diag.panel = panel.hist)
```

>>>>

```{r load data}
# Ecological data: leaf counts, infection
load("~/GitHub/superspreaders/stems_plots.RData")
# Vegetation data
veg_us_2005 <- read.csv("analysis/data/understory_veg_2005.csv")
veg_us_2011 <- read.csv("analysis/data/understory_veg_2011.csv")
# Untagged dbh data
untagged_dbh <- read.csv("analysis/data/untagged_dbh_summary.csv")

# Physical environment: topography
library(rgdal)
plots_shp <- readOGR("analysis/data/plot_gis/", "soco_plots_metric")
summary(plots_shp)
plot(plots_shp, axes=T)
plots_env <- plots_shp@data

# Temperature: rainy season hour summaries
library(plyr); library(dplyr)
plots_temps <- read.csv("analysis/data/rainy_season_temps.csv")
summary(plots_temps)
plots_temps <- select(plots_temps, -X)
```

Join the plot level environmental data (geographic/topographic & temperatures)
```{r join plot data, message=FALSE}
plots_env <- left_join(plots_env, plots_temps, by = c("PLOT_ID"="plotid"))
summary(plots_env)
str(plots_env)
plots_env <- plots_env %>% select(-cat) %>% rename(easting = X, northing = Y, plotid = PLOT_ID)
```

Investigating `NAs` in the `plots_env` data set
```{r investigate NAs}
summary(plots_env)
filter(plots_env, is.na(sample_year))

@@

Examine the bay laurel data at individual and plot levels
```{r examine umca data}
summary(plots_umca)
# This only has number of infected, uninfected, and total, which must apply to living stems only. I want to analyze leaf counts.
summary(bay_laurel)
# This has stem-level data on bay laurel.
```

I can use this to calculate plot level leaf count variables (sum, average) for each plot and year
```{r calculate plot level leaf counts}
plots_umca <- left_join(plots_umca, bay_laurel %>% filter(status == "Alive") %>% 
      group_by(plot, year) %>% 
      summarise(
            avg_lfct = mean(slc, na.rm = T),
            tot_lfct = sum(slc, na.rm = T)),
      by = c("plot", "year"))
summary(plots_umca)
```

```{r examine lfct NAs}
filter(plots_umca, is.na(avg_lfct))
# I'm not sure why this wasn't excluded from the calculations with the na.rm = T option
filter(bay_laurel, plot == "YAHNG02", status == "Alive")
# I'm removing it manually from the `plots_umca` data frame
plots_umca <- plots_umca %>% filter(avg_lfct != "NaN")
summary(plots_umca)
```

So now I have a plot-level of UMCA infection levels for each year. I want to join this with some of the environmental data.

```{r join env and umca data}
summary(plots_env)
plots_env$avg_psi <- rowMeans(cbind(plots_env$psiSpr, plots_env$psiSum, plots_env$psiWin))
summary(plots_umca)
# plot == plotid, year == sample_year
plots_umca <- left_join(select(plots_umca, -species), plots_env, by = c("plot" = "plotid", "year" = "sample_year"))
summary(plots_umca)
filter(plots_umca, is.na(easting))
# NAs are due to no microclimate data for year 2003, and abandoned plots BUSH01 & SUGAR02
```

Now I also want to calculate some plot-level community variables from the vegetation data. First, understory richness from the vegetation transects.
```{r calculate understory richness}
summary(veg_us_2005)
veg_us_2005 <- as.tbl(select(veg_us_2005, -X))
summary(veg_us_2011)
veg_us_2011 <- as.tbl(select(veg_us_2011, -X))
veg_us_2011
veg_us_2005

veg <- rbind(veg_us_2005, veg_us_2011)
veg
unique(veg$year)
veg$year[veg$year < 2011] <- 2005 # Creates the two most complete veg sampling
veg <- veg %>% rename(us_species = UnderstorySp, plotid = PlotID)
richness <- veg %>% group_by(plotid, year) %>% summarise(us_rich = length(unique(us_species)))
richness
```

Next, overstory richness from the DBH data for tagged & untagged stems.

```{r calculate overstory richness}
summary(untagged_dbh)
untagged_dbh <- as.tbl(untagged_dbh %>% select(-X) %>% rename(plotid = PlotID, species = Species))
untagged_dbh
unique(untagged_dbh$year)
untagged_dbh$year[untagged_dbh$year < 2014] <- 2005 # Create two most complete groups
untagged_dbh$plotid <- tolower(untagged_dbh$plotid)

untagged_dbh %>% group_by(plotid, year) %>% summarise(untagged_rich = length(unique(species)))

stems <- as.tbl(stems)
summary(stems)
unique(stems$status)
stems <- stems %>% select(-X) %>% filter(status == "Alive" | status == "Dead")
unique(stems$species)
stems$species <- tolower(stems$species)
stems$species[stems$species == "quke x quag"] <- "quag x quke"
stems$plotid <- tolower(stems$plotid)
stems %>% group_by(plotid, year) %>% summarise(tagged_rich = length(unique(species)))

os_richness <- left_join(
      stems %>% group_by(plotid, year) %>% summarise(tagged_rich = length(unique(species))),
      untagged_dbh %>% group_by(plotid, year) %>% summarise(untagged_rich = length(unique(species))), by = c("plotid", "year")
) %>% mutate(os_rich = tagged_rich + untagged_rich)
os_richness
richness <- left_join(os_richness, richness, by = c("plotid","year"))
richness
summary(richness)
```

I also want to calculate some variables related to DBH, and in particular the "importance value" for each species. Importance value is calculated in the literature as:

`IV = 0.5 * (relative density / relative dominance)` where 
`Relative Density = # stems of species / total # of stems` and
`Relative Dominance = basal area of species / total basal area`

```{r calculate importance value}
unique(untagged_dbh$species)
untagged_dbh
unique(stems$species)

tagged_dbh <- left_join(stems %>% 
      filter(status == "Alive", location == "In") %>%
      select(plotid, year, species, dbh) %>% 
      group_by(plotid, species, year) %>% 
      summarise(live_count = length(species), live_avg_dbh = mean(dbh), live_tot_dbh = sum(dbh)),
      stems %>% 
      filter(status == "Dead", location == "In") %>%
      select(plotid, year, species, dbh) %>% 
      group_by(plotid, species, year) %>% 
      summarise(dead_count = length(species), dead_avg_dbh = mean(dbh), dead_tot_dbh = sum(dbh)),
      by = c("plotid", "species", "year")
)
tagged_dbh <- as.tbl(tagged_dbh)
summary(tagged_dbh)

tagged_dbh
untagged_dbh

stems_dbh <- rbind(tagged_dbh, untagged_dbh)

stems_dbh <- stems_dbh %>% group_by(plotid, year) %>% mutate(all_live_count = sum(live_count), all_live_dbh = sum(live_tot_dbh, na.rm = T))
summary(stems_dbh)
filter(stems_dbh, all_live_dbh > 800)

stems_dbh <- stems_dbh %>% mutate(rel_dens = live_count / all_live_count, rel_dom = live_tot_dbh / all_live_dbh, imp_value = 0.5*(rel_dens + rel_dom))
summary(stems_dbh)
```

Make a pairs plot of the continuous variables for UMCA plots.
```{r pairs plot}
# Helper functions for `pairs`
panel.hist <- function(x, ...)
{
      usr <- par("usr"); on.exit(par(usr))
      par(usr = c(usr[1:2], 0, 1.5) )
      h <- hist(x, plot = FALSE)
      breaks <- h$breaks; nB <- length(breaks)
      y <- h$counts; y <- y/max(y)
      rect(breaks[-nB], 0, breaks[-1], y, col = "cyan", ...)
}
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
      usr <- par("usr"); on.exit(par(usr))
      par(usr = c(0, 1, 0, 1))
      r <- abs(cor(x, y, use = "na.or.complete"))
      txt <- format(c(r, 0.123456789), digits = digits)[1]
      txt <- paste0(prefix, txt)
      if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
      text(0.5, 0.5, txt, cex = cex.cor * r)
}

plots_umca <- ungroup(plots_umca) %>% select(-starts_with("psi"), -plot, -year)

pairs(select(plots_umca, -plot, -year, -starts_with("psi")), lower.panel = panel.cor, diag.panel = panel.hist)
```
